A generalized definition of multidimensional item response theory parameters

### 2022-09-14 output/paper_draft/paper_M2PLM_parameters.Rmd
Credits Ackerman (1994, 2005) for acknowledging that Reckase (1985) and Reckase & McKinley (1991) imply orthogonal axes

A crucial assumption leading to those formulations is that the axes of the
latent space are orthogonal
[see @reckase_difficulty_1985, p. 404 and Equation 7].
Of course, orthogonality is a property of a geometrical space instead of a
latent trait space.
Thus, nothing prevents a researcher from interpreting an abstraction such as
a latent trait space, no matter its structure, as an orthogonal space
--exploratory factor analysis software such as SPSS <!--# Reference needed? -->or
the R package *psych* <!--# Reference needed? -->graphically represent
factor loadings on orthogonal axes, even after an oblique rotation.
Nevertheless, the assimilation of correlated latent traits to non-orthogonal
axes is prevalent in the common factor literature
[see, e.g., @harman_modern_1970].<!--# Thurstone's Multiple Factor Analysis, 1947? -->
This assumption is not so easily found in the MIRT literature though,
which mostly deals with uncorrelated dimensions
[see @reckase_multidimensional_2009 for a comprehensive treatment on the topic].
This fact is probably due to the different spotlights
the two theoretical programs put the focus on;
Factor analysis aims at a clear interpretation of
the latent covariance structure, and thus focuses on finding
a structure matrix as simple as possible. MIRT literature, on the other hand,
aims at modeling responses to certain stimuli and estimating person scores,
being the model interpretation secondary in relevance.<!--# Reference needed? -->
Nevertheless, the confluence of the two theoretical approaches into a common
modern measurement theory [@mcdonald_test_1999] imposes
the relaxation of traditional constrains on both approaches.
It is not uncommon to find recent examples of MIRT models where the
correlation matrix of the latent dimensions is freely estimated
[see, e.g., @reckase_multidimensional_2009, pp. 224-228]. Despite this, the
correspondence between correlated latent traits and orthogonal coordinate axes
is usually implicit in the MIRT literature without further consideration<!--# Exaples? -->
[see @zhang_theoretical_1999 for a notable exception].
What is more, disregarding this geometrical interpretation may lead to
misconceptions that put the measurement and its conclusions into question.<!-- completar?? -->

@@

We must make a few remarks here. First, as we are not constraining ourselves to
cognitive traits, we speak of _positive_ instead of _correct response_.
This may imply a correct response, if we are talking about a cognitive test, or
endorsing an item (or a certain response option) in a non-cognitive domain.
Second, although not explicitly stated, $`r DISCR_VECTOR`$ is a vector of
non-negative real numbers in Reckse
[-@reckase_difficulty_1985, a decrease in the probability of a correct response with an increase in ability makes little sense].
In our case, $`r DISCR_VECTOR`$ is a vector of real numbers; the value
of the $`r DIM_INDEX`$-th component $`r DISCR_PARAM`$ can be negative if
the item is *inverse* for (i.e., taps the negative pole of)
latent dimension $`r DIM_INDEX`$.
Finally, as responses are not necessarily interpreted as correct or incorrect,
parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_,
instead of the item difficulty.

>>>>

A crucial assumption leading to those formulations is that the axes of the
latent space are orthogonal
[@ackerman_using_1994; @ackerman_multidimensional_2005; see also @reckase_difficulty_1985, p. 404 and Equation 7].
Of course, orthogonality is a property of a geometrical space instead of a
latent trait space.
Thus, nothing prevents a researcher from interpreting an abstraction such as
a latent trait space, no matter its structure, as an orthogonal space
--exploratory factor analysis software such as SPSS <!--# Reference needed? -->or
the R package *psych* <!--# Reference needed? -->graphically represent
factor loadings on orthogonal axes, even after an oblique rotation.
Nevertheless, the assimilation of correlated latent traits to non-orthogonal
axes is prevalent in the common factor literature
[see, e.g., @harman_modern_1970].<!--# Thurstone's Multiple Factor Analysis, 1947? -->
This assumption is not so easily found in the MIRT literature though,
which mostly deals with uncorrelated dimensions
[see @reckase_multidimensional_2009 for a comprehensive treatment on the topic].
This fact is probably due to the different spotlights
the two theoretical programs put the focus on;
Factor analysis aims at a clear interpretation of
the latent covariance structure, and thus focuses on finding
a structure matrix as simple as possible. MIRT literature, on the other hand,
aims at modeling responses to certain stimuli and estimating person scores,
being the model interpretation secondary in relevance.<!--# Reference needed? -->
Nevertheless, the confluence of the two theoretical approaches into a common
modern measurement theory [@mcdonald_test_1999] imposes
the relaxation of traditional constrains on both approaches.
It is not uncommon to find recent examples of MIRT models where the
correlation matrix of the latent dimensions is freely estimated
[see, e.g., @reckase_multidimensional_2009, pp. 224-228]. Despite this, the
correspondence between correlated latent traits and orthogonal coordinate axes
is usually implicit in the MIRT literature without further consideration<!--# Exaples? -->
[see @zhang_theoretical_1999 for a notable exception].
What is more, disregarding this geometrical interpretation may lead to
misconceptions that put the measurement and its conclusions into question.<!-- completar?? -->

@@

We must make a few remarks here. First, as we are not constraining ourselves to
cognitive traits, we speak of _positive_ instead of _correct response_.
This may imply a correct response, if we are talking about a cognitive test, or
endorsing an item (or a certain response option) in a non-cognitive domain.
Second, although not explicitly stated, $`r DISCR_VECTOR`$ is a vector of
non-negative real numbers in Reckase
[-@reckase_difficulty_1985, a decrease in the probability of a correct response with an increase in ability makes little sense].
In our case, $`r DISCR_VECTOR`$ is a vector of real numbers; the value
of the $`r DIM_INDEX`$-th component $`r DISCR_PARAM`$ can be negative if
the item is *inverse* for (i.e., taps the negative pole of)
latent dimension $`r DIM_INDEX`$.
Finally, as responses are not necessarily interpreted as correct or incorrect,
parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_,
instead of the item difficulty.
